### 2023-02-07 chapters/background.tex
adding fairXplainer

\subsection{Global Sensitivity Analysis: Variance Decomposition}
Global sensitivity analysis is a field that studies how the global uncertainty in the output of a function can be attributed to the different sources of uncertainties in the input while considering the whole input domain~\cite{saltelli2008global}.
Sensitivity analysis is an essential component for quality assurance and impact assessment of models in EU~\cite{eu}, USA~\cite{usepa}, and research communities~\cite{saltelli2020five}.
%\paragraph{Sensitivity Analysis:}
\emph{Variance-based sensitivity analysis} is a form of global sensitivity analysis, where variance is considered as the measure of uncertainty~\cite{sobol1990sensitivity,sobol2001global}. To illustrate, let us consider a real-valued function $  \fiffunc(\mathbf{Z}) $, where $ \mathbf{Z} $ is a vector of $ \numnonsensitive $ input variables $ \{Z_1, \dots, Z_k\} $. Now, we decompose $ \fiffunc(\mathbf{Z}) $ among the subsets of inputs, such that:

\begin{align}
	\fiffunc(\mathbf{Z}) &= \fiffunc_0 + \sum_{i=1}^{\numnonsensitive} \fiffunc_{\{i\}}(Z_i) +  \sum_{i < j}^{\numnonsensitive} \fiffunc_{\{i,j\}}(Z_i, Z_j)  + \cdots  + \fiffunc_{\{1, 2, \dots, \numnonsensitive\}} (Z_1, Z_2, \dots, Z_{\numnonsensitive})\notag%\label{eq:functional_decomposition}
	\\    
	&= \fiffunc_0 +  \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} \fiffunc_{\mathbf{S}}(\mathbf{Z}_{\mathbf{S}})\label{eq:functional_decomposition_set_notation}
\end{align}

In this decomposition, $ \fiffunc_0 $ is a constant, $ \fiffunc_{\{i\}} $ is a function of $ Z_i $, $ \fiffunc_{\{i,j\}} $ is a function of $ Z_i $ and $ Z_j $, and so on. Here, $ [\numnonsensitive] \triangleq \{1,2,\dots, \numnonsensitive\} $ and $ \mathbf{S}$ is an ordered subset of $[\numnonsensitive] \setminus \emptyset $.  We denote $ \mathbf{Z}_{\mathbf{S}}  \triangleq \{Z_i | i \in \mathbf{S}\}$ as the input of $ \fiffunc_{\mathbf{S}} $, where $ \mathbf{Z}_{\mathbf{S}}$ is a set of variables with indices belonging to $ \mathbf{S} $.  The standard condition of this decomposition is the orthogonality of each term in the right-hand side of Eq.~\eqref{eq:functional_decomposition_set_notation}~\cite{sobol1990sensitivity}. $ \fiffunc_{\mathbf{S}}(\mathbf{Z}_{\mathbf{S}}) $ is the effect of varying all the features in $\mathbf{Z}_{\mathbf{S}}$ simultaneously. %\unsure{Is this sentence necessary?}  
For $|\mathbf{S}|=1$, it quantifies an individual variable's effect. For $|\mathbf{S}|>1$, it quantifies the higher-order interactive effect of variables.

Now, if we assume $ \fiffunc $ to be square integrable, we obtain the decomposition of the variance of the output~\cite{sobol1990sensitivity}.

\begin{align}\label{eq:variance_decomposition_set_notation}
	\mathsf{Var}[g(\mathbf{Z})] &= \sum_{i=1}^{\numnonsensitive}V_{\{i\}} +  \sum_{i<j}^{\numnonsensitive} V_{\{i,j\}}  + \cdots  + V_{\{1, 2\dots, \numnonsensitive\}}= \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} V_{\mathbf{S}} 
\end{align}

where $ V_{\{i\}} $ is the variance of $ \fiffunc_{\{i\}} $, $ V_{\{i,j\}} $ is the variance of $ \fiffunc_{\{i,j\}} $ and so on. Formally,

\[ V_{\mathbf{S}} \triangleq \mathsf{Var} _{\mathbf{Z}_{\mathbf{S}}}\left[\mathbb{E}_{{\textbf {Z}}/\mathbf{Z}_{\mathbf{S}}}[g(\mathbf{Z})\mid \mathbf{Z}_{\mathbf{S}}]\right]- \sum_{\mathbf{S}' \subset \mathbf{S}/\emptyset}{V} _{\mathbf{S}'}.\]

Here, $\mathbf{S}'$ denotes all the ordered proper subsets of $\mathbf{S}$. %\improvement{Check notations.}
This variance decomposition shows how the variance of $\fiffunc(\mathbf{Z}) $ can be decomposed into terms attributable to each input, as well as the interactive effects among them. Together all terms sum to the total variance of the model output.

>>>>

\subsection{Global Sensitivity Analysis (GSA): Variance Decomposition}
Global sensitivity analysis is an active field of research that studies how the global uncertainty in the output of a function can be attributed to the different sources of uncertainties in the input variables~\cite{saltelli2008global}.
Sensitivity analysis is an essential component for quality assurance and impact assessment of models in EU~\cite{eu}, USA~\cite{usepa}, and research communities~\cite{saltelli2020five}.
%\paragraph{Sensitivity Analysis:}
\emph{Variance-based sensitivity analysis} is a form of global sensitivity analysis, where variance is considered as the measure of uncertainty~\cite{sobol1990sensitivity,sobol2001global}. To illustrate, let us consider a real-valued function $  \fiffunc(\feature) $, where $ \feature $ is a vector of $ \numfeatures $ input variables $ \{Z_1, \dots, Z_\numfeatures\} $.
% and $ Y $ is a univariate output. 
%\blue{In sensitivity analysis, assumptions are made on the independence and uniform distribution of $ X_i $'s within the unit hypercube}, that is, $ X_i \in [0,1] $ for $ i = \{ 1, \dots, k\} $.\footnote{In our experiments, we have not considered this yet.} Let $ [\numnonsensitive] \triangleq \{1,2,\dots, \numnonsensitive\} $. Then, 
Now, we decompose $ \fiffunc(\feature) $ among the subsets of inputs, such that:
\begin{align}
	\fiffunc(\feature) &= \fiffunc_0 + \sum_{i=1}^{\numfeatures} \fiffunc_{\{i\}}(Z_i) +  \sum_{i < j}^{\numfeatures} \fiffunc_{\{i,j\}}(Z_i, Z_j)  + \cdots\notag + \fiffunc_{\{1, 2, \dots, \numfeatures\}} (Z_1, Z_2, \dots, Z_{\numfeatures})\\
	&= \fiffunc_0 +  \sum_{\mathbf{S} \subseteq [\numfeatures]} \fiffunc_{\mathbf{S}}(\feature_{\mathbf{S}})\label{fairness_fairXplainer_eq:functional_decomposition_set_notation}
\end{align}
%	$S_> = \{{S_0}, \ldots, {S_{|s|}}\}$ i.e. a monotonically increasing ordering of $|S|$ features for a given starting feature $Z_{S_0}$ such that $S_0 <S_1 <\ldots <S_{|S|}$.
The standard condition of this decomposition is the orthogonality of each term in the right-hand side of Eq.~\eqref{fairness_fairXplainer_eq:functional_decomposition_set_notation}~\cite{sobol1990sensitivity}. In this decomposition, $ \fiffunc_0 $ is a constant, $ \fiffunc_{\{i\}} $ is a function of $ Z_i $, $ \fiffunc_{\{i,j\}} $ is a function of $ Z_i $ and $ Z_j $, and so on. Adhering to the set-based notations, we denote by $ g_{\mathbf{S}} $ a function of a non-empty subset of variables $ \feature_{\mathbf{S}}  \triangleq \{Z_i \mid i \in \mathbf{S}\} \subseteq \feature $, where $ \mathbf{S} = \{S_i \mid 1 \le |S_i| \le \numfeatures\} \subseteq [\numfeatures] $ is a non-empty subset of indices with $ [\numfeatures] \triangleq \{1,2,\dots, \numfeatures\}  $.  Here, $|\mathbf{S}|=1$ quantifies an individual variable's effect while $|\mathbf{S}|>1$ quantifies the higher-order intersectional effect of variables.

Considering $ \fiffunc $ as square integrable, we obtain the decomposition of the variance of $ \fiffunc(\feature) $ expressed as the sum of variances of $\fiffunc_{\mathbf{S}}$'s~\cite{sobol1990sensitivity}.
\begin{align}\label{fairness_fairXplainer_eq:variance_decomposition_set_notation}
	\mathsf{Var}[g(\feature)] &= \sum_{i=1}^{\numfeatures}V_{\{i\}} +  \sum_{i<j}^{\numfeatures} V_{\{i,j\}}  + \dots +  V_{\{1, 2\dots, \numfeatures\}}= \sum_{\mathbf{S} \subseteq [\numnonsensitive]} V_{\mathbf{S}} 
\end{align}
where $ V_{\{i\}} $ is the variance of $ \fiffunc_{\{i\}} $, $ V_{\{i,j\}} $ is the variance of $ \fiffunc_{\{i,j\}} $ and so on. Formally,   
\begin{equation}
	\label{eq:decomposed_variance}
	V_{\mathbf{S}} \triangleq \mathsf{Var} _{\feature_{\mathbf{S}}}\left[\mathbb{E}_{{\textbf {Z}}\setminus\feature_{\mathbf{S}}}[g(\feature)\mid \feature_{\mathbf{S}}]\right]- \sum_{\mathbf{S}' \subset \mathbf{S}}{V} _{\mathbf{S}'}.
\end{equation} 
Here, $\mathbf{S}'$ denotes all the non-empty and ordered proper subsets of $\mathbf{S}$. Thus, $ V_{\mathbf{S}} $ is the variance w.r.t.\ $ \feature_{\mathbf{S}} $ by subtracting the variances of all non-empty proper subsets of $ \feature_{\mathbf{S}} $. As a result, Eq.~\eqref{fairness_fairXplainer_eq:variance_decomposition_set_notation} demonstrates how the variance of $\fiffunc(\feature) $ can be decomposed into terms attributable to each input feature, as well as the intersectional effects among them. Conversely, together all terms sum to the total variance of the model output.

### 2023-02-07 chapters/fairness/fif/abstract.tex
adding fairXplainer

\begin{abstract}

	Fairness in machine learning has attained significant focus due to the widespread application of machine learning in high-stake decision-making tasks. Unless regulated with a fairness objective, machine learning classifiers might demonstrate unfairness/bias towards certain demographic populations in the data. Thus, the quantification and mitigation of the bias induced by classifiers have become a central concern. In this paper, \textit{we aim to quantify the influence of different features on the bias of a classifier}. To this end, we propose a framework of \emph{Fairness Influence Function} (FIF), and compute it as a scaled difference of conditional variances in the classifier's prediction. We also instantiate an algorithm, {\fairXplainer}, that uses variance decomposition among the subset of features and a local regressor to compute FIFs accurately, while also capturing the intersectional effects of the features. Our experimental analysis validates that {\fairXplainer} captures the influences of both individual features and higher-order feature interactions, estimates the bias more accurately than existing local explanation methods, and detects the increase/decrease in bias due to affirmative/punitive actions in the classifier.

\end{abstract}

>>>>

\begin{abstract}
	Fairness in machine learning has attained significant focus due to the widespread application in high-stake decision-making tasks. Unregulated machine learning classifiers can exhibit bias towards certain demographic groups in data, thus the quantification and mitigation of classifier bias is a central concern in fairness in machine learning.  In this paper, \textit{we aim to quantify the influence of different features in a dataset on the bias of a classifier}. To do this, we introduce the \textit{Fairness Influence Function} (FIF). This function breaks down bias into its components among individual features and the intersection of multiple features. The key idea is to represent existing group fairness metrics as the difference of the scaled conditional variances in the classifier's prediction and apply a decomposition of variance according to global sensitivity  analysis. To estimate FIFs, we instantiate an algorithm {\framework} that applies variance decomposition of classifier's prediction following local regression. 	Experiments demonstrate that {\framework} captures FIFs of individual feature and intersectional features, provides a better approximation of bias based on FIFs, demonstrates higher correlation of FIFs with fairness interventions, and detects changes in bias due to fairness affirmative/punitive actions in the classifier.
\end{abstract}
